---
title: "Lab 3: Adding Layers with ggplot2 and Data Preparation"
author: "Muhammad Ali"
subtitle: "Visualizing `mtcars`, Wrangling Tidy Data, and Simple Predictions"
format: html
execute:
  echo: true
  warning: false
  message: false
---

## Learning goals

By the end of this lab, you should be able to:

1.  Prepare and recode variables using modern `dplyr` pipelines (`%>%` and `|>`).
2.  Build a `ggplot2` chart one layer at a time, including faceting.
3.  Build and label a scatter plot that compares different car metrics.

## 1. Load the Tidyverse and Inspect the Data

We will use the built-in `mtcars` dataset. First, load `tidyverse`, which includes both `dplyr` (for data manipulation) and `ggplot2` (for visualization).

```{r}
library(tidyverse)

head(mtcars)
```

We will focus on:

-   `wt`: weight, in 1,000 pounds
-   `mpg`: miles per gallon
-   `am`: transmission type (`0` = automatic, `1` = manual)
-   `cyl`: number of cylinders (`4`, `6`, or `8`)

## 2. Data Management: Cleaning and Labeling Variables

Instead of transforming data types inside our plotting code, good data-management practice means preparing our dataset explicitly first. We will convert numeric codes into informative factors.

This lab uses `%>%` for its `dplyr` pipelines. The base R pipe, `|>`, moves data from the left side into the function on the right in the same general way.

```{r}
mtcars_clean <- mtcars %>%
  mutate(
    transmission = factor(am, levels = c(0, 1), labels = c("Automatic", "Manual")),
    cylinders = factor(cyl)
  )

glimpse(mtcars_clean)
```

## 3. Build a Plot Layer by Layer

### Step 3a: Initialize the Canvas

`ggplot()` defines the dataset and basic visual mapping using `aes()`.

```{r}
ggplot(mtcars_clean, aes(x = wt, y = mpg))
```

### Step 3b: Add Data Points and Color

We use `+` to add `geom_point()`. Mapping `color = transmission` uses our newly cleaned factor variable.

```{r}
ggplot(mtcars_clean, aes(x = wt, y = mpg, color = transmission)) +
  geom_point(size = 3)
```

### Step 3c: Add Trend Lines and Labels

`geom_smooth()` adds linear trend lines (`method = "lm"`). `labs()` updates the title and axis labels.

```{r}
ggplot(mtcars_clean, aes(x = wt, y = mpg, color = transmission)) +
  geom_point(size = 3) +
  geom_smooth(method = "lm", se = FALSE) +
  labs(
    title = "Fuel Efficiency vs. Car Weight",
    subtitle = "Comparing Automatic and Manual Transmissions",
    x = "Weight (1,000 lbs)",
    y = "Miles Per Gallon (MPG)",
    color = "Transmission Type"
  ) +
  theme_minimal()
```

## 4. Multi-Panel Visualizations (Faceting)

To avoid visual clutter when comparing subgroups, we can split plots into side-by-side panels using `facet_wrap()`.

```{r}
ggplot(mtcars_clean, aes(x = wt, y = mpg, color = transmission)) +
  geom_point(size = 3) +
  geom_smooth(method = "lm", se = FALSE) +
  facet_wrap(~ cylinders) +
  labs(
    title = "Fuel Efficiency vs. Weight Faceted by Engine Cylinders",
    x = "Weight (1,000 lbs)",
    y = "Miles Per Gallon (MPG)",
    color = "Transmission"
  ) +
  theme_minimal()
```

## 5. Practice: Build and Label Your Own Plot

Now create your own plot using `mtcars_clean`. Choose two numeric car metrics to compare. You may use:

-   `mpg`: miles per gallon
-   `wt`: weight, in 1,000 pounds
-   `hp`: horsepower
-   `disp`: engine displacement
-   `qsec`: time to complete one-quarter mile

Add a point for each car and label each point with its number of cylinders. Put `label = cylinders` inside `aes()` in `geom_text()`. Use `labs()` to give your chart a clear title, axis labels, and a legend label for the number of cylinders.

```{r}
# Choose two numeric variables for x and y.
# Add geom_point() and geom_text(aes(label = cylinders)).
# Use labs() to explain your chart clearly.

ggplot(mtcars_clean, aes(x = hp , y = mpg, color = transmission)) +
  geom_point(size = 3) +
  geom_smooth(method = "lm", se = FALSE) +
  facet_wrap(~ cylinders) +
  labs(
    title = "Fuel Efficiency vs Horsepower Faceted by Engine Cylinders",
    x = "Horsepower",
    y = "Miles Per Gallon (MPG)",
    color = "Transmission"
  ) +
  theme_bw()
```

## Reflection Questions

1.  Which two car metrics did you compare, and what pattern do you see in your plot?
2.  How do the cylinder labels help you interpret the points?
3.  What layer did you add last, and what did it change in the chart?

1\) The car metrics I compared were horsepower and miles per gallon. The patterns I see within my plot is that both manual and automatic cars seem to have a negative trend line as the cylinders go up.

2\) The cylinder labels help me interpret my points by showing different trends within different cylinder groups. As seen in the previous plot, 2 out of the three faceted plots had different trendlines whereas this plot they all seem to be the same trend lines for the 3 plots.

3\) I put `theme_bw()` as it shows a different aesthetic for the 3 plots than `theme_minimal()` aesthetic. It shows the engine cylinder in a gray box at the top to indicate the categorical number and then the the separated plot as mentioned.
