---
title: "Week3B - Window Functions"
author: "Muhammad Ali"
format:
  html:
    code-fold: false
    
execute:
  echo: true
  warning: false
  message: false

editor: visual
---

# Window Functions Analysis

## Approach

### Goal

The goal of this assignment is to utilize and understand the concept of <mark>Window Functions </mark> used in R or SQL.

### Window Functions

What is a window function? It preserves information from the previous row as it iterates through the next row.

Some functions include, average, sum or any type of function that involves grouping things.

### The Plan

The first thing to do is to generate a dataset, synthetic or based off real data, with an LLM or find a dataset on popular sites such as:

-   Kaggle

-   Google Databases

-   UCI Machine Learning Repository

Following that dataset, we need to do the following steps:

-   Clean the dataset (if necessary)

-   Understand window functions in SQL or R

-   Apply the concept of window functions for the dataset

### Dataset

Given that this assignment allows leniency with either searching for a dataset online or generating a synthetic dataset using an LLM, I have taken the LLM approach and asked Gemeni to generate a synthetic dataset.

-   [Github Dataset CSV](https://raw.githubusercontent.com/HaiderrX/CUNY-SPS-MSDS/refs/heads/main/DATA607/Week%203%20-%20R%20Character%20Manipulation%20and%20Date%20Processing/inventory_synthetic_data%20-%20Sheet1.csv)

## Codebase

Before being able to do anything with this csv file, the first thing to do was separate all of the data in Microsoft Excel using the `=TEXTSPLIT()` function. The reason this is essential is because when using a generated dataset from an LLM it pastes all of the information in one column separated by columns. So having to separate them into new columns using that function was a must. Following that we will need to import that dataset into R and to get rid of the first column as it is being used as a reference column.

```{r libraries}
library(DBI)
library(RPostgres)
library(tidyverse)
library(dplyr)
library(readr)
```

### Data Cleaning & Transformation

```{r importing-dataset}
raw_data <- read_csv("https://raw.githubusercontent.com/HaiderrX/CUNY-SPS-MSDS/refs/heads/main/DATA607/Week%203%20-%20R%20Character%20Manipulation%20and%20Date%20Processing/inventory_synthetic_data%20-%20Sheet1.csv")

head(raw_data)
colnames(raw_data)
```

As mentioned, we need to get rid of the first column in order to proceed. However we can also rename columns for conventional reasons.

```{r data-cleaning}
clean_data <- raw_data |>
  select(
    date = Date,
    product_id = Product_ID,
    product_name = Product_Name,
    units_sold = Units_Sold,
    price_per_unit = Price_Per_Unit,
    revenue = Revenue,
    ending_inventory_count = Ending_Inventory_Count
  )
```

```{r}
head(clean_data)
glimpse(clean_data)
```

Data has been cleaned, we can now proceed.

### SQL Connection

```{r SQL-connector}
connection <- dbConnect(
  RPostgres::Postgres(),
  dbname = "products_db",
  host = "localhost",
  port = 5432,
  user = "postgres",
  password = ""
)
```

```{r creating-table}

dbExecute(connection, "DROP TABLE IF EXISTS products_info;")

create_table_query <- "
CREATE TABLE IF NOT EXISTS products_info (
    date DATE,
    product_id VARCHAR(50),
    product_name TEXT,
    units_sold INT,
    prices_per_unit NUMERIC(10, 2),
    revenue NUMERIC(10,2),
    ending_inventory_count INT
);
"

dbExecute(connection, create_table_query)
```

```{r write-to-sql-table}
dbWriteTable(
  con = connection,
  name = "products_info",
  value = clean_data,
  overwrite = TRUE,
  field.types = c(
    date                   = "DATE",
    product_id             = "VARCHAR(50)",
    product_name           = "TEXT",
    units_sold             = "INTEGER",
    price_per_unit         = "NUMERIC(10, 2)",
    revenue                = "NUMERIC(10, 2)",
    ending_inventory_count = "INTEGER"
  )
)


# dbGetQuery(connection, "SELECT * FROM products_info")
# Table works and everything is listed
```

Successfully imported the table to SQL, now to start Window Function queries.

### Window Functions

#### Average Year To date

In this query, we want to get the average year to date on the average number of units sold for a product.

```{r average-year-to-date}
products_avg <- dbGetQuery(connection, "
SELECT date,
      product_id,
      product_name,
      units_sold,
      revenue,
      ROUND (
        AVG(units_sold) OVER (
          PARTITION BY product_id
          ORDER BY date
        )::numeric, 2 
      ) AS avg_units_sold
FROM products_info;
") |>
  as_tibble()

products_avg
```

#### The Six Day Moving Average for Each Item

In this query, we want the six day moving average for each item, which means we want to know what the average of something was for a six day period

```{r six-day-moving-avg}
moving_avg <- dbGetQuery(connection, "
SELECT
	product_id,
	product_name,
	date,
	ROUND(
		AVG(units_sold) OVER (
	        PARTITION BY product_id 
	        ORDER BY date 
	        ROWS BETWEEN 5 PRECEDING AND CURRENT ROW
		)::numeric, 2
	) AS six_day_moving_avg
FROM products_info;                         
") |>
  as_tibble()

moving_avg
```

```{r closing sql connection}
dbDisconnect(connection)
```

## Conclusion

Working with SQL is very enjoyable for me as it is one of my favorite languages to work with as I can easily translate queries to the English language similarly to Python.

With that being said, the frustrating part of this was the SQL connection and being able to write the rows of the CSV to SQL so I had to research how to do it which led me to the `dbWriteTable()` function which made it easier for me to import the dataframe to SQL.

I would love to work with SQL queries in the future, and maybe more problems to do them with so I can get better at articulating the queries and the steps it takes to do them in.
