---
title: "DATA 607: Week 3"
subtitle: "Recommender Systems, Window Functions, and Character Data"
author: "Darwhin Gomez"
format:
  html:
    output-file: week3_character_strings_page.html
    toc: true
    code-fold: true

editor: visual
execute:
  echo: true
---

## DATA 607

### Week 3: Character Data and String Manipulation

::: notes
Tonight we will work with text as data. Our examples use Jane Austen novels, but the same tools work for names, survey responses, addresses, product titles, and log files.
:::

## Tonight's agenda

-   Due this week and next week
-   Global baseline estimates for recommenders
-   Window functions in SQL
-   Why character data needs cleaning
-   Working with strings in R
-   `janeaustenr` examples
-   A first look at regular expressions
-   Practice and next steps

## Due this week

+---------------+----------------------------------------------------------+
| Thursday      | Assignment 3A Global Baseline Estimate: Approach \|      |
+---------------+----------------------------------------------------------+
| Thursday      | Assignment 3B Window Functions: Approach                 |
+---------------+----------------------------------------------------------+
| Sunday        | Quiz: Data Centric AI                                    |
+---------------+----------------------------------------------------------+
| Sunday        | Discussion: Data Centric AI                              |
+---------------+----------------------------------------------------------+
| Sunday        | Assignment 3A Global Baseline Estimate: Code Base        |
+---------------+----------------------------------------------------------+
| Sunday        | Assignment 3B Window Functions: Code Base                |
+---------------+----------------------------------------------------------+
| Monday        | Assignment 3A Global Baseline Estimate: Video Explainer  |
+---------------+----------------------------------------------------------+
| Monday        | Assignment 3B Global Baseline Estimate: Video Explainer\ |
+---------------+----------------------------------------------------------+

## Due next week

-   Thursday: Project 1 approach
-   Sunday: Project 1 code base
-   Sunday: Scenario Design quiz and discussion
-   Monday: Project 1 video explainer

## Global baseline estimates

A recommender system estimates how a user might rate an item they have not yet rated.

The simplest useful prediction starts with a baseline:

-   the typical rating across the whole dataset
-   how a particular critic tends to rate
-   how a particular item tends to be rated

## The bias-based formula

$$
\widehat{r}_{ui} = \mu + b_u + b_i
$$

| Term  | Meaning                                                      |
|-------|--------------------------------------------------------------|
| $\mu$ | Global mean rating                                           |
| $b_u$ | User bias: this user rates above or below average            |
| $b_i$ | Item bias: this item receives above or below average ratings |

The hat means "predicted," not an observed rating.

## Critic-movie ratings data {.smaller}

Your ratings matrix has one row per critic and one column per movie. Blank cells mean a rating was not observed; the question mark is the rating we want to predict.

```{r}
# Load dplyr and tidyr functions for this example.
library(tidyverse)

ratings_wide <- tribble(
  ~critic, ~CaptainAmerica, ~Deadpool, ~Frozen, ~JungleBook, ~PitchPerfect2, ~StarWarsForce,
  "Burton", NA, NA, NA, 4, NA, 4,
  "Charley", 4, 5, 4, 3, 2, 3,
  "Dan", NA, 5, NA, NA, NA, 5,
  "Dieudonne", 5, 4, NA, NA, NA, 5,
  "Matt", 4, NA, 2, NA, 2, 5,
  "Mauricio", 4, NA, 3, 3, 4, NA,
  "Max", 4, 4, 4, 2, 2, 4,
  "Nathan", NA, NA, NA, NA, NA, 4,
  "Param", 4, 4, 1, NA, NA, 5,
  "Parshu", 4, 3, 5, 5, 2, 3,
  "Prashanth", 5, 5, 5, 5, NA, 4,
  "Shipra", NA, NA, 4, 5, NA, 3,
  "Sreejaya", 5, 5, 5, 4, 4, 5,
  "Steve", 4, NA, NA, NA, NA, 4,
  "Vuthy", 4, 5, 3, 3, 3, NA,
  "Xingjia", NA, NA, 5, 5, NA, NA
)
ratings_wide

ratings <- ratings_wide |>
  pivot_longer(-critic, names_to = "movie", values_to = "rating") |>
  drop_na(rating)

ratings
```

`pivot_longer()` changes the ratings matrix into one row per observed critic-movie rating.

## Step 1: global mean

```{r}
global_mean <- mean(ratings$rating)
global_mean
```

If we know nothing else, this is our prediction for every user-movie pair.

## Step 2a: critic bias

A critic bias measures whether that critic usually rates above or below the global mean.

```{r}
critic_bias <- ratings |>
  group_by(critic) |>
  summarize(b_u = mean(rating) - global_mean)

critic_bias
```

A positive value means the critic tends to give higher ratings than the typical rating.

## Step 2b: movie bias

A movie bias measures whether a movie usually receives ratings above or below the global mean.

```{r}
movie_bias <- ratings |>
  group_by(movie) |>
  summarize(b_i = mean(rating) - global_mean)

movie_bias
```

A negative value means the movie tends to receive lower ratings than the typical rating.

## Predict Param's `PitchPerfect2` rating

Do the arithmetic by hand before writing any code.

+-----------------------------+------------------------------+---------------+
| Quantity                    | Calculation                  | Rounded value |
+=============================+==============================+==============:+
| Global mean, $\mu$          | mean of all observed ratings | 3.93          |
+-----------------------------+------------------------------+---------------+
| Param bias, $b_u$           | $3.50 - 3.93$                | -0.43         |
+-----------------------------+------------------------------+---------------+
| `PitchPerfect2` bias, $b_i$ | $2.71 - 3.93$                | -1.22         |
+-----------------------------+------------------------------+---------------+

**Prediction:** `3.93 + (-0.43) + (-1.22) = 2.28`

Your task is to create the tibble of biases and implement this arithmetic. Then score every movie Param has not rated and recommend the highest predicted rating.

## Recommend a movie for Param

First, identify movies Param has not rated. Then calculate the baseline prediction for every candidate and sort from highest to lowest.

```{r}
param_seen <- ratings |>
  filter(critic == "Param") |>
  distinct(movie)
param_seen

candidate_predictions <- ratings |>
  distinct(movie) |>
  anti_join(param_seen, by = "movie") |>
  mutate(critic = "Param") |>
  left_join(critic_bias, by = "critic") |>
  left_join(movie_bias, by = "movie") |>
  mutate(predicted_rating = global_mean + b_u + b_i) |>
  arrange(desc(predicted_rating))

candidate_predictions
```

Recommend the first row: `JungleBook`. It has the highest global-baseline prediction among the movies Param has not rated.

## When is a global baseline useful?

**Useful for**

-   a transparent first model
-   sparse data or cold-start comparisons
-   checking a more complex recommender

**Not enough for**

-   user-item interactions
-   content similarity
-   changing preferences over time

## Window functions: PostgreSQL to R

We will work with the `afc_east` table in the `NFL607` PostgreSQL database.

1.  Connect to PostgreSQL from R.
2.  Import the table into `afc_east_df`.
3.  Send SQL window-function queries to PostgreSQL.
4.  Bring each query result back to R as a tibble.

The rendered examples use the local NFL607 database so you can inspect each returned result.

## Connect to the NFL607 database

```{r}
#| eval: true
# Install once if needed:
# install.packages(c("DBI", "RPostgres"))

library(DBI)
library(RPostgres)
library(tidyverse)

pg_pw <- Sys.getenv("NFL607_DB_PASSWORD")
if (!nzchar(pg_pw)) stop("Set NFL607_DB_PASSWORD in .Renviron before rendering.")

con <- dbConnect(
  RPostgres::Postgres(),
  dbname = "NFL607",
  host = "localhost",
  port = 5432,
  user = "postgres",
  password = pg_pw
)
```

Store the password in an environment variable. Do not put a real password in a Quarto file or Git repository.

## Import the SQL table into an R data frame

```{r}
#| eval: true
afc_east_df <- dbReadTable(con, "afc_east") |>
  as_tibble() |>
  mutate(game_date = as.Date(game_date))

head(afc_east_df)
```

`afc_east_df` is now an R tibble. The PostgreSQL table remains unchanged in the database.

## Window functions keep every game row

`GROUP BY` reduces many game rows to one row per team. A window function adds a calculation while preserving every game row.

``` sql
AVG(home_points) OVER (PARTITION BY home_team)
```

Read the query result directly into a tibble with `dbGetQuery()`.

## Overall average and differential

```{r}
#| eval: true
overall_scores_df <- dbGetQuery(con, "
  SELECT home_team,
         game_date,
         home_points,
         ROUND(AVG(home_points) OVER (), 1) AS home_avg,
         ROUND(home_points - AVG(home_points) OVER (), 1) AS home_differential
  FROM afc_east
  ORDER BY home_team, game_date
") |>
  as_tibble()

overall_scores_df
```

`OVER ()` means every row belongs to one window: the full result set.

## Average by home team

```{r}
#| eval: true
team_average_df <- dbGetQuery(con, "
  SELECT home_team,
         game_date,
         home_points,
         ROUND(AVG(home_points) OVER (PARTITION BY home_team), 1) AS home_team_avg
  FROM afc_east
  ORDER BY home_team, game_date
") |>
  as_tibble()

team_average_df
```

`PARTITION BY home_team` restarts the calculation for each team.

## Rank scores within each team

```{r}
#| eval: true
ranked_games_df <- dbGetQuery(con, "
  SELECT home_team,
         game_date,
         home_points,
         RANK() OVER (
           PARTITION BY home_team
           ORDER BY home_points DESC
         ) AS home_rank
  FROM afc_east
  ORDER BY home_team, home_rank
") |>
  as_tibble()

ranked_games_df
```

The highest home score for each team receives rank 1.

## Running total by team

```{r}
#| eval: true
running_scores_df <- dbGetQuery(con, "
  SELECT home_team,
         game_date,
         home_points,
         SUM(home_points) OVER (
           PARTITION BY home_team
           ORDER BY game_date
           ROWS BETWEEN UNBOUNDED PRECEDING AND CURRENT ROW
         ) AS cumulative_home_points
  FROM afc_east
  ORDER BY home_team, game_date
") |>
  as_tibble()

running_scores_df
```

The window starts with each team's first game and ends at the current game row.

## The R equivalent after import

After `afc_east_df` is in R, `dplyr` can perform a similar running-total calculation.

```{r}
#| eval: true
afc_east_df |>
  arrange(home_team, game_date) |>
  group_by(home_team) |>
  mutate(cumulative_home_points = cumsum(home_points)) |>
  ungroup()
```

The SQL version calculates in PostgreSQL. The R version calculates after the table has been imported.

## PostgreSQL window functions mapped to dplyr

A PostgreSQL window function is usually a `dplyr` `mutate()` after the data has been grouped and ordered. Both approaches add a value to every original row.

+-----------------------------+----------------------+----------------------------------------+----------------------------------------------------------+
| PostgreSQL idea             | PostgreSQL pattern   | `dplyr` counterpart                    | What it does                                             |
+=============================+======================+========================================+==========================================================+
| All rows in one window      | `AVG(x) OVER ()`     | `mutate(avg_x = mean(x))`              | Repeats one overall value on every row.                  |
+-----------------------------+----------------------+----------------------------------------+----------------------------------------------------------+
| A window for each group     | `PARTITION BY team`  | `group_by(team)`                       | Restarts a calculation for each team.                    |
+-----------------------------+----------------------+----------------------------------------+----------------------------------------------------------+
| Order inside a window       | `ORDER BY game_date` | `arrange(game_date, .by_group = TRUE)` | Sets the order used by ranks, lags, and running totals.  |
+-----------------------------+----------------------+----------------------------------------+----------------------------------------------------------+
| Keep every original row     | `... OVER (...)`     | `mutate(...)`                          | Adds a calculated column without reducing rows.          |
+-----------------------------+----------------------+----------------------------------------+----------------------------------------------------------+
| Reduce to one row per group | `GROUP BY team`      | `group_by(team) |> summarise(...)`     | Produces a summary table instead of a row-by-row result. |
+-----------------------------+----------------------+----------------------------------------+----------------------------------------------------------+

## Aggregate functions: PostgreSQL to dplyr

+-------------------------------------------+-------------------------------------------------+------------------------------+
| PostgreSQL                                | `dplyr` inside `mutate()`                       | `dplyr` inside `summarise()` |
+===========================================+=================================================+==============================+
| `AVG(x) OVER (...)`                       | `mean(x, na.rm = TRUE)`                         | `mean(x, na.rm = TRUE)`      |
+-------------------------------------------+-------------------------------------------------+------------------------------+
| `SUM(x) OVER (...)`                       | `sum(x, na.rm = TRUE)`                          | `sum(x, na.rm = TRUE)`       |
+-------------------------------------------+-------------------------------------------------+------------------------------+
| `MIN(x) OVER (...)` / `MAX(x) OVER (...)` | `min(x, na.rm = TRUE)` / `max(x, na.rm = TRUE)` | Same functions               |
+-------------------------------------------+-------------------------------------------------+------------------------------+
| `COUNT(*) OVER (...)`                     | `n()`                                           | `n()`                        |
+-------------------------------------------+-------------------------------------------------+------------------------------+

Use these functions inside `mutate()` when you want to keep each game row. Use them inside `summarise()` when you want one result per group.

## Ranks, offsets, and running totals: PostgreSQL to dplyr

+----------------------------------------------------------------------+---------------------------------+-----------------------------------------------------------------+
| PostgreSQL                                                           | `dplyr` counterpart             | Meaning                                                         |
+======================================================================+=================================+=================================================================+
| `RANK() OVER (ORDER BY x DESC)`                                      | `min_rank(desc(x))`             | Equal values share a rank, and later ranks have gaps.           |
+----------------------------------------------------------------------+---------------------------------+-----------------------------------------------------------------+
| `DENSE_RANK() OVER (ORDER BY x DESC)`                                | `dense_rank(desc(x))`           | Equal values share a rank, with no gaps afterward.              |
+----------------------------------------------------------------------+---------------------------------+-----------------------------------------------------------------+
| `ROW_NUMBER() OVER (ORDER BY x DESC)`                                | `row_number(desc(x))`           | Gives every row a unique position, including tied values.       |
+----------------------------------------------------------------------+---------------------------------+-----------------------------------------------------------------+
| `LAG(x) OVER (ORDER BY date)`                                        | `lag(x)` after `arrange(date)`  | Reads the previous row's value.                                 |
+----------------------------------------------------------------------+---------------------------------+-----------------------------------------------------------------+
| `LEAD(x) OVER (ORDER BY date)`                                       | `lead(x)` after `arrange(date)` | Reads the next row's value.                                     |
+----------------------------------------------------------------------+---------------------------------+-----------------------------------------------------------------+
| `SUM(x) OVER (... ROWS BETWEEN UNBOUNDED PRECEDING AND CURRENT ROW)` | `cumsum(x)` after `arrange()`   | Adds values from the first ordered row through the current row. |
+----------------------------------------------------------------------+---------------------------------+-----------------------------------------------------------------+

## One dplyr pipeline that matches the SQL examples

```{r}
#| eval: true
window_map_df <- afc_east_df |>
  group_by(home_team) |>
  arrange(game_date, .by_group = TRUE) |>
  mutate(
    home_team_avg = mean(home_points, na.rm = TRUE),
    home_rank = min_rank(desc(home_points)),
    prior_home_points = lag(home_points),
    next_home_points = lead(home_points),
    cumulative_home_points = cumsum(home_points)
  ) |>
  ungroup()

window_map_df
```

Read this pipeline from top to bottom: `group_by()` matches `PARTITION BY`; `arrange()` matches `ORDER BY`; and `mutate()` matches the `OVER (...)` calculation that keeps each game row.

## When SQL and dplyr differ

Use PostgreSQL when the data is stored in the database and you want the database to do the work. Use `dplyr` after importing data when you need to continue analyzing it in R. In both tools, choose `mutate()`/`OVER()` to keep rows and `summarise()`/`GROUP BY` to collapse rows.

## Close the database connection

```{r}
#| eval: true
dbDisconnect(con)
```

Close the connection when you are done querying the database.

## Window-function checklist

Before writing the query, state in words:

1.  What does one row represent?

2.  Which rows should be compared together?

3.  What order matters?

4.  Should the output preserve every original row?

    For example: "For each team, ordered by game date, calculate home points scored to date." \## Why character data matters

## Text is everywhere in data work:

-   names, emails, and addresses
-   survey comments and reviews
-   categories imported from files
-   dates stored as text
-   IDs such as `NY-2026-0142`

## Before analysis, text often needs to be standardized, searched, split, or combined.

## Learning goals

By the end of this session, you should be able to:

-   recognize a character vector in R
-   clean inconsistent capitalization and whitespace
-   find text with `str_detect()`
-   extract or replace part of a string
-   use simple regular expressions to describe a pattern

## Packages for tonight

```{r}
# Install once if needed:
# install.packages(c("tidyverse", "janeaustenr"))

library(tidyverse)
library(janeaustenr)
```

`stringr` is included when you load the tidyverse.

## What is a string?

A **string** is text inside quotation marks.

```{r}
movie <- "Pride and Prejudice"
year <- "1813"
rating <- "4.5"
```

All three objects are character strings, even when their contents look like a number.

## Character vectors

A character vector stores multiple strings.

```{r}
characters <- c("Elizabeth", "Darcy", "Jane", "Bingley")
characters

class(characters)
length(characters)
```

Each position is one string; the whole object is one vector.

## Load a novel as data

`janeaustenr` provides complete Austen novels as character vectors.

```{r}
pride <- prideprejudice

length(pride)
head(pride, 8)
```

Each element of `pride` is one line from *Pride and Prejudice*.

## Make a small tibble

Tibbles work especially well with `dplyr` and `stringr`.

```{r}
pride_tbl <- tibble(
  line_number = seq_along(pride),
  text = pride
)

pride_tbl |> slice(1:6)
```

## Count characters

`str_length()` counts the characters in each string.

```{r}
str_length(c("Darcy", "Elizabeth", "Mr. Bingley"))

pride_tbl |>
  mutate(characters = str_length(text)) |>
  arrange(desc(characters)) |>
  slice(1:3)
```

## Change capitalization

Use a predictable case before matching names or categories.

```{r}
names <- c("ELIZABETH", "mr. darcy", "JaNe")

str_to_lower(names)
str_to_upper(names)
str_to_title(names)
```

Do not change case if capitalization carries meaning in your data.

## Remove extra whitespace

Whitespace can make identical-looking values fail to match.

```{r}
messy_names <- c("  Elizabeth", "Darcy  ", "  Jane  Bennet  ")

str_trim(messy_names)      # remove edges
str_squish(messy_names)    # remove edges and repeat spaces
```

## Replace text

`str_replace()` changes the first match. `str_replace_all()` changes every match.

```{r}
line <- "Mr. Darcy spoke to Mr. Bingley."

str_replace(line, "Mr\\.", "Mister")
str_replace_all(line, "Mr\\.", "Mister")
```

The double backslash tells R to send a literal backslash to the pattern engine.

## Combine strings

`str_c()` joins strings together.

```{r}
first <- c("Elizabeth", "Fitzwilliam")
last  <- c("Bennet", "Darcy")

str_c(first, last, sep = " ")
str_c("Chapter ", 1:3)
```

`sep` is placed between pieces. `collapse` joins a whole vector into one string.

## Find a word

`str_detect()` returns `TRUE` or `FALSE` for each string.

```{r}
str_detect(
  c("Elizabeth smiled.", "Darcy was silent.", "Jane arrived."),
  "Elizabeth"
)
```

This is useful because `filter()` keeps rows where a condition is `TRUE`.

## Search the novel

Find lines that mention Elizabeth.

```{r}
pride_tbl |>
  filter(str_detect(text, "Elizabeth")) |>
  slice(1:5)
```

This search is case-sensitive: `"elizabeth"` would not match `"Elizabeth"`.

## Ignore capitalization in a search

Wrap the pattern in `regex(..., ignore_case = TRUE)`.

```{r}
pride_tbl |>
  filter(str_detect(text, regex("darcy", ignore_case = TRUE))) |>
  slice(1:5)
```

Use this when you want `Darcy`, `DARCY`, and `darcy` to count as the same text.

## Count matches

`str_count()` counts how often a pattern appears in each string.

```{r}
str_count(
  c("ha", "ha ha", "perhaps"),
  "ha"
)
```

Unlike `str_detect()`, this returns a number rather than `TRUE` or `FALSE`.

## Which lines mention Darcy most?

```{r}
pride_tbl |>
  mutate(darcy_mentions = str_count(text, "Darcy")) |>
  filter(darcy_mentions > 0) |>
  arrange(desc(darcy_mentions)) |>
  select(line_number, darcy_mentions, text) |>
  slice(1:5)
```

This keeps the original line and adds a new measurement.

## Extract part of a string

`str_sub()` selects characters by position.

```{r}
books <- c("Emma", "Persuasion", "Mansfield Park")

str_sub(books, 1, 3)
str_sub(books, -4, -1)
```

Negative positions count backward from the end of a string.

## Split into columns

When every value has the same structure, `separate_wider_delim()` is often clearer.

```{r}
tibble(code = c("NY-2026-0142", "CA-2026-0143")) |>
  separate_wider_delim(
    code,
    delim = "-",
    names = c("state", "year", "record_id")
  )
```

This converts one packed character column into three usable columns.

## Useful regex building blocks

| Pattern | Meaning           | Example match          |
|---------|-------------------|------------------------|
| `.`     | any one character | `c.t` matches `cat`    |
| `\\d`   | a digit           | `2026` contains digits |
| `\\s`   | whitespace        | a space or tab         |
| `^`     | start of a string | `^Mr`                  |
| `$`     | end of a string   | `ing$`                 |

Use regex one small piece at a time and test your pattern.

## Practice: find a pattern

Use `str_detect()` to find IDs that begin with `NY-`.

```{r}
ids <- c("NY-2026-0142", "CA-2026-0143", "NY-2025-0088")

# Write your pattern here:
str_detect(ids, "...")
```

Hint: the start-of-string symbol is `^`.

## A dependable text-cleaning workflow

1.  Inspect the original strings.
2.  Decide what consistent means for this task.
3.  Add a cleaned column; preserve the original.
4.  Test a few known examples.
5.  Count or filter to check the result.
6.  Document any rule that removes or changes data.

## Key functions to remember

| Goal             | Function                           |
|------------------|------------------------------------|
| Count characters | `str_length()`                     |
| Clean spaces     | `str_trim()`, `str_squish()`       |
| Change case      | `str_to_lower()`, `str_to_title()` |
| Find text        | `str_detect()`                     |
| Count matches    | `str_count()`                      |
| Replace text     | `str_replace_all(), str_replace()` |
| Extract text     | `str_sub()`                        |
| Split text       | `str_split()`                      |

## 
